pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-03-02 bmc_article.tex

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.png}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, and normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates. \textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling. The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. An exception was \textit{scVI} (Supplementary Figure 12), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section). Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

\begin{figure}
    \centering
    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only. Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average MI of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 13). We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 14).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 15A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 15B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 16-18). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 16-17). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.65]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterisk denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 18).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 12). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

\begin{figure}
    \centering
    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr_files/figure-html/dimred-1.png}}
    \caption{\textbf{Evaluation of common dimensionality reduction methods. A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions. \textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
    \label{fig:figure7}
\end{figure}

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 13 and 19). This tended to affect the accuracy of the clustering (Supplementary Figure 20), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A). Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 21) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An implication is that the commonly-used elbow method most likely underestimates dimensionality.

\begin{figure}
    \centering
    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality_files/figure-html/figure6-1.png}}
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization. \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
    \label{fig:figure8}
\end{figure}

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 21), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 22), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:figure9}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

\begin{figure}
    \centering
    \includegraphics[scale=0.41, keepaspectratio]{{main_figures/clustering_files/figure-html/clustering-1.png}}
    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
    \label{fig:figure9}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.49, keepaspectratio]{{main_figures/imputation_files/figure-html/imputation-1.png}}
     \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
    \label{fig:figure10}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.08,keepaspectratio]{main_figures/pipeComp_summary.png}
    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
    \label{fig:figure11}
\end{figure}

@@

The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming. As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method. The resulting evaluation is available in the Supplementary Figure 12.

@@

% \section*{Figures}
%   \begin{figure}[h!]
%   \caption{\csentence{Sample figure title.}
%       A short description of the figure content
%       should go here.}
%       \end{figure}
% 
% \begin{figure}[h!]
%   \caption{\csentence{Sample figure title.}
%       Figure legend text.}
%       \end{figure}

>>>>

\iffalse

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.pdf}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets.pdf}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering.pdf}}
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}
\fi

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. An exception was \textit{scVI} (Supplementary Figure 13), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section). Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization.pdf}}
    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. 'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. }
    \label{fig:figure5}
\end{figure}
\fi

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14). We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 17A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterisk denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}
\fi

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 13). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr.pdf}}
    \caption{\textbf{Evaluation of common dimensionality reduction methods. A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions. \textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
    \label{fig:figure7}
\end{figure}
\fi

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 14 and 21). This tended to affect the accuracy of the clustering (Supplementary Figure 22), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:figure8}A). Figure \ref{fig:figure8}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:figure8}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:figure8}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:figure8}C-D), when combined with \textit{sctransform} or the standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:figure8}C) and resulted in the best clustering accuracy (Figure \ref{fig:figure8}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An implication is that the commonly-used elbow method most likely underestimates dimensionality.

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality.pdf}}
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization. \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
    \label{fig:figure8}
\end{figure}
\fi

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:figure9}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 1 and 23), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 24), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:figure9}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.34, keepaspectratio]{{main_figures/clustering.pdf}}
    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
    \label{fig:figure9}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.49, keepaspectratio]{{main_figures/imputation.pdf}}
     \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
    \label{fig:figure10}
\end{figure}
\fi

@@

\iffalse
\begin{figure}[h]
    \centering
    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
    \label{fig:figure11}
\end{figure}
\fi

@@

The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming. As it was not practically feasible to perform the training step on many combination of parameters as it was done with other tools, we evaluated it on a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method. The resulting evaluation is available in the Supplementary Figure 13.

@@

\section*{Figures}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Overview of the benchmark datasets used.}}
    \label{fig:figure1}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. 'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. }
    \label{fig:figure5}
\end{figure}
